---
title: "Scientific subscription costs in Finland 2010-2015"
author: [Leo Lahti](http://www.iki.fi/Leo.Lahti)
date: "`r Sys.Date()`"
---


```{r foi-init, message=FALSE, echo=FALSE}
library(dplyr)
library(tidyr)
library(reshape2)
library(ggplot2)
theme_set(theme_bw(20))
# License fee data URL
f <- "~/data/Tietopyynto/20160607\ Kustantajahintatiedot_koottu_v0.99.csv"
# Read the data
df <- read.csv(f, fileEncoding = "latin1")
# Rename
df = rename(df, Publisher = Kustantaja.Välittäjä)
df = rename(df, Organization = Organisaation.nimi)
df = rename(df, Material = Aineistotyyppi)
df = rename(df, Type = Hankintatapa)
df = rename(df, Year = Vuosi)
df = rename(df, Price = Hinta)
```


Subscription costs 2010-2015: the Finnish case
===========

### Background

Following [Tim Gowers successful FOI request on scientific
subscription costs in
UK](http://gowers.wordpress.com/2014/04/24/elsevier-journals-some-facts/),
we realized by the [OKF Finland](http://fi.okfn.org/wg/openscience/)
and [other open science
enthusiasts](https://www.facebook.com/groups/241398182642057/permalink/411482855633588)
that a similar FOI request could be made in Finland. I left the FOI
request to Finnish universities in summer 2014. After their denial and
esoteric explanations we appealed to court. The positive decision was
given in August 2015. Then it took another year to collect and tidy up
all the information across the Finnish universities and other
institutions. The process is summarized in another [blog post](). The
Finnish Ministry of Education [Open Science
Initiative](http://openscience.fi) then collected and organized the
data, which has now been released on [their website]().

Whereas related data have been collected in a few countries (see
Stuart Lawson's [recent
post](http://stuartlawson.org/2016/06/publicly-available-data-on-international-journal-subscription-costs),
Finland is one of the very first countries (after US and UK) where the
subscription costs are available for individual publishers over a
several years time span. Here, I present preliminary analysis of this
open data set. The fully reproducible source code of this document is
[available](https://github.com/antagomir/temp/blob/master/20160610/foi.Rmd).



### Overall costs 2010-2015

```{r foi-totalcosts, message=FALSE, echo=FALSE, fig.show="hold", out.width="430px", fig.width = 8, fig.height = 5}
dfs = df %>% group_by(Organisaation.tyyppi) %>% summarise(Costs = sum(Price)/1e6) %>% arrange(Costs)
dfs$Organisaation.tyyppi = factor(dfs$Organisaation.tyyppi, levels = as.character(dfs$Organisaation.tyyppi))
p <- ggplot(dfs, aes(x = Organisaation.tyyppi, y = Costs)) + geom_bar(stat = "identity") + ggtitle("Costs by organization type") + coord_flip() + xlab("") + ylab("Total Costs (MEUR)")
print(p)

dfs = df %>% group_by(Organization) %>% summarise(Costs = sum(Price)/1e6) %>% arrange(desc(Costs))
dfs$Organization = factor(dfs$Organization, levels = rev(as.character(dfs$Organization)))
p <- ggplot(dfs[1:20,], aes(x = Organization, y = Costs)) + geom_bar(stat = "identity") + ggtitle("Costs by organization") + coord_flip() + xlab("") + ylab("Total Costs (MEUR)")
print(p)
names(dfs) = c("Organization", "Total (MEUR)")
write.csv(dfs, file = "table/cost_by_organization.csv", quote = F, row.names = F)

dfs = df %>% group_by(Publisher) %>% summarise(Costs = sum(Price)/1e6) %>% arrange(desc(Costs))
dfs$Publisher = factor(dfs$Publisher, levels = rev(as.character(dfs$Publisher)))
p <- ggplot(dfs[1:20,], aes(x = Publisher, y = Costs)) + geom_bar(stat = "identity") + ggtitle("Costs by publisher") + coord_flip() + xlab("") + ylab("Total Costs (MEUR)")
print(p)
names(dfs) = c("Publisher", "Total (MEUR)")
write.csv(dfs, file = "table/cost_by_publisher.csv", quote = F, row.names = F)


dfs = df %>% group_by(Material) %>% summarise(Costs = sum(Price)/1e6) %>% arrange(desc(Costs))
dfs$Material = factor(dfs$Material, levels = rev(as.character(dfs$Material)))
p <- ggplot(dfs[1:20,], aes(x = Material, y = Costs)) + geom_bar(stat = "identity") + ggtitle("Costs by material") + coord_flip() + xlab("") + ylab("Total Costs (MEUR)")
print(p)
names(dfs) = c("Material", "Total (MEUR)")
write.csv(dfs, file = "table/cost_by_material.csv", quote = F, row.names = F)

dfs = df %>% group_by(Type) %>% summarise(Costs = sum(Price)/1e6) %>% arrange(desc(Costs))
dfs$Type = factor(dfs$Type, levels = rev(as.character(dfs$Type)))
names(dfs) = c("Type", "Total (MEUR)")
write.csv(dfs, file = "table/cost_by_type.csv", quote = F, row.names = F)

dfs = df %>% group_by(Year) %>% summarise(Costs = sum(Price)/1e6) %>% arrange(desc(Year))
names(dfs) = c("Year", "Total (MEUR)")
write.csv(dfs, file = "table/cost_by_year.csv", quote = F, row.names = F)
```

### Total costs for Finland 2010-2015

The total costs paid to scientific publishers by Finland 2010-2015. The annual increase is indicated.

```{r foi-timeline, message=FALSE, eval=TRUE, echo=FALSE, fig.show="hold", out.width="400px"}
# Price, Year : all publishers
# Growth percentage compared to previous year is shown
dfs = df %>% group_by(Year) %>% summarise(Costs = sum(Price)/1e6) 
p <- ggplot(dfs, aes(x = Year, y = Costs)) +
       geom_bar(stat = "identity") + ggtitle("Total costs for Finland") +
       geom_text(data = dfs[-1,], aes(x = Year, y = 1.042 * Costs,
     label = paste(round(100 * diff(dfs$Costs)/dfs$Costs[-length(dfs$Costs)], 1), "%", sep = "")   ),
     	   size = 8) +
       scale_x_continuous(breaks = 2010:2015, labels = c(2010:2015)) + 
       ylab("Costs (MEUR)")
print(p)
```


## Publication costs with top publishers 2010-2015

 * [Total costs by organization (complete table)](table/cost_by_organization.csv)
 * [Total costs by publisher (complete table)](table/cost_by_publisher.csv)
 * [Total costs by material](table/cost_by_material.csv)
 * [Total costs by type](table/cost_by_type.csv)
 * [Total costs by year](table/cost_by_year.csv)   

```{r foi-toppublishercosts, message=FALSE, eval=TRUE, echo=FALSE, fig.show="hold", out.width="200px"}
# Total cost : all publishers
dfs <- df %>% group_by(Publisher) %>% summarise(Cost = sum(Price)/1e6) %>% arrange(desc(Cost))
colnames(dfs) = c("Publisher", "Cost (MEUR)")
kable(head(dfs, 20))
```


## Costs by publisher over time

Top-10 publishers (out of `r nrow(dfs)`) correspond to `r round(100 * sum(dfs[["Cost (MEUR)"]][1:10])/sum(dfs[["Cost (MEUR)"]]))`% of the overall costs.

```{r foi-timebypublisher, message=FALSE, echo=FALSE, fig.width=10, fig.height=5}
# Price, Year : individual publishers compared
# Growth percentage compared to previous year is shown
top.publishers <- as.character(dfs$Publisher[1:10])
dfs <- dplyr::filter(df, Publisher %in% top.publishers) %>% group_by(Year, Publisher) %>% summarise(Costs = sum(Price)/1e6) %>% arrange(Year)
dfs$Publisher <- factor(dfs$Publisher, levels = top.publishers)
p <- ggplot(dfs, aes(x = Year, y = Costs, color = Publisher)) +
       geom_point() +
       geom_line() +       
       ggtitle("Price development by publisher") +
       ylab("Price (MEUR)") 
print(p)
```


