### Relative increase in costs 2015 vs 2010

```{r foi-time, message=FALSE, eval=TRUE}
dfs <- df %>% group_by(Year, Publisher) %>% summarise(Price = sum(Price)/1e6) %>% arrange(Year)
dfss <- spread(dfs, Year, Price)
kasvu <- unlist(dfss[, "2015"]/dfss[, "2010"]);
names(kasvu) <- as.character(dfss$Publisher)
#rev(sort(kasvu))

# Relative prices with 2010 baseline
# Top relative increase
kustantajat <- dfss[,1]
hinnat = as.matrix(dfss[, -1])
hinnat <- hinnat/hinnat[,1]
dfs2 <- as.data.frame(hinnat)
dfs2$Publisher <- as.character(unlist(kustantajat, use.names = F))
dfs2 <- dfs2[!is.na(dfs2[, "2015"]),]
dfs2 <- dfs2[rev(order(dfs2[, "2015"])),]
top <- as.character(unlist(dfs2$Publisher, use.names = F)[1:10])
dfs3 <- dfs[unlist(dfs$Publisher) %in% top,]
dfs3$Publisher <- as.character(unlist(dfs3$Publisher, use.names = F))
dfs3$Publisher <- factor(dfs3$Publisher, levels = top)
dfs3$Year <- as.numeric(as.character(dfs3$Year))
dfs3$Price <- as.numeric(as.character(dfs3$Price))
p <- ggplot(dfs3,
       aes(x = Year, y = Price, color = Publisher)) +
       geom_point() +
       geom_line() +       
       ggtitle("Relative price development by publisher") +
       ylab("Price (EUR)") 
print(p)
```


```{r foi-test, message=FALSE, echo=FALSE, fig.width=10, fig.height=5}
URL <- paste0(
        "https://cdn.rawgit.com/christophergandrud/networkD3/",
        "master/JSONdata/energy.json")
Energy <- jsonlite::fromJSON(URL)
# Plot
sankeyNetwork(Links = Energy$links, Nodes = Energy$nodes, Source = "source",
             Target = "target", Value = "value", NodeID = "name",
             units = "TWh", fontSize = 12, nodeWidth = 30)
```

dfs = df %>% group_by(Publisher) %>% summarise(Costs = sum(Price)/1e6) %>% arrange(desc(Costs))
dfs$Publisher = factor(dfs$Publisher, levels = rev(as.character(dfs$Publisher)))
dfs$Percentage = round(100 * dfs$Costs/sum(dfs$Costs), 1)
names(dfs) = c("Publisher", "Total (MEUR)", "Percent (%)")
write.csv(dfs, file = "table/cost_by_publisher.csv", quote = F, row.names = F)

```{r}
library(DT)
DT::datatable(iris)
```


df$Aineisto

library(stringr)
e = list()
s = str_trim(as.character(df$Aineisto))
#e$nodes = data.frame(name = unique(unlist(strsplit(as.character(df$Aineisto), ";"))))
e$nodes = data.frame(name = s)
e$links

